Learning action models with minimal observability

### 2019-05-01 doc/aij18/minor-revision/evaluation.tex
AIJ18 minor revision folder created

In this section we introduce the metrics used by \FAMA to evaluate the action models that result from solving a learning task $\Lambda$. \textcolor{red}{First, we will motivate the use of two standard syntactic metrics ({\em precision} and {\em recall}). Then, section \ref{syntactic_precision_recall} will define these metrics for planning models, and section \ref{semantic_precision_recall} will introduce a semantic evaluation measure that builds upon {\em precision} and {\em recall}}. Finally, section \ref{edit_distance} explains how \FAMA computes our novel semantic-based metrics.

\textcolor{red}{We can observe in Table \ref{table:models_comparison2} that most of the approaches use a similar syntax-based metric that consists in (1) counting the missing and extra fluents that appear in the learned model wrt the GTM and (2) normalizing this error by the the total number of all the possible preconditions and effects of an action model. This is an \emph{optimistic} metric since error rates are not normalized by the size of the actual GTM. The set of preconditions and effects of the GTM is usually smaller than the set of all possible preconditions and effects and thereby it turns out that these syntax-based metrics may output error rates below 100\% for totally wrong learned models. To overcome this limitation we propose to use two standard metrics from ML, {\em precision} and {\em recall}, that are frequently used in pattern recognition, information retrieval and binary classification~\cite{davis2006relationship}. These two syntactic metrics are generally more informative than counting the number of errors between the learned action models and the GTM:}

%When the planning reference model of the input observations (i.e., the GTM) is available, the quality of the learned action models is measurable using two well-studied syntax-based metrics, {\em precision} and {\em recall}, commonly used in tasks such as information retrieval and recommender systems~\cite{davis2006relationship}. These two syntactic metrics are generally more informative than counting the number of errors between the learned action models and the GTM. Intuitively, precision gives a notion of {\em soundness} while recall gives a notion of the {\em completeness} of the learned models:

@@

\begin{mydefinition}
	Let $PRE(\xi) = \underset{p \in pre(\xi)}{\bigcup} pre_p(\xi)$, $ADD(\xi) = \underset{p \in add(\xi)}{\bigcup} add_p(\xi)$, and $DEL(\xi) = \underset{p \in del(\xi)}{\bigcup} del_p(\xi)$ be the set of propositional fluents that represent preconditions, positive and negative effects of a given action model $\xi$. We define:
	\begin{small}
		\textcolor{red}{
			\begin{align*}
			INS(\mathcal{M}, \mathcal{M'})=&\bigcup\limits_{\xi\in\mathcal{M}, \xi'\in\mathcal{M'} s.t.\ name(\xi) = name(\xi')}
			PRE(\xi') \backslash PRE(\xi) \cup
			ADD(\xi') \backslash ADD(\xi) \cup
			DEL(\xi') \backslash DEL(\xi)
			\\
			\\
			DEL(\mathcal{M}, \mathcal{M'})=&\bigcup\limits_{\xi\in\mathcal{M}, \xi'\in\mathcal{M'} s.t.\ name(\xi) = name(\xi')}
			PRE(\xi) \backslash PRE(\xi') \cup
			ADD(\xi) \backslash ADD(\xi') \cup
			DEL(\xi) \backslash DEL(\xi')
			\end{align*}
		}
	\end{small}
\end{mydefinition}

With these ingredients in mind, we adapt the definitions of syntactic precision and recall to domain models. Let $\mathcal{M}$ be a domain model and let $\mathcal{M'}$ be the GTM. We know that \textcolor{red}{$size(\mathcal{M}) = \sum_{\xi \in \mathcal{M}}\left|pre(\xi)\right| + \left|add(\xi)\right| + \left|del(\xi)\right|$} and by definition the number of preconditions and effects of the learned action models is equal to the sum of {\em true positives} and {\em false positives}; that is, $size(\mathcal{M}) = tp + fp$.

@@

\begin{mydefinition} \label{syn-precision} The precision of $\mathcal{M}$ relative to the GTM is defined as the fraction of the common preconditions and effects between $\mathcal{M}$ and the GTM among all prediconditions and effects of $\mathcal{M}$.
%This gives an intuitive measure of the soundness of $\mathcal{M}$.
	\begin{small}
		\begin{align*}
		Precision=\frac{tp}{tp+fp}=\frac{size(\mathcal{M})- \left|DEL(\mathcal{M},GTM)\right|}{size(\mathcal{M})}
		\end{align*}
	\end{small}
\end{mydefinition}

\begin{mydefinition} \label{syn-recall} The recall of $\mathcal{M}$ relative to the GTM is defined as the fraction of the common preconditions and effects between $\mathcal{M}$ and the GTM among all preconditions and effects of the GTM.
%This gives an intuitive measure of the completeness of $\mathcal{M}$.

@@

\textcolor{red}{Intuitively, precision gives a notion of {\em soundness} while recall gives a notion of the {\em completeness} of the learned models. We interpret a sound learned model as one in which all preconditions and effects are correct with respect to the GTM, and so there is no need to remove anything. A complete model is one in which no precondition or effect is missing.
%In more detail, precision is the fraction of relevant predicates among the predicates that make up the learned model, and recall is the fraction of relevant predicates in the learned model over all relevant predicates, understanding a relevant predicate as one that appears in the GTM.
As an example, a precision of 0.5 means that only half of the predicates that make up the learned domain model are present in the GTM, while a recall of 0.5 means that only half of the predicates that make up the GTM are present in the learned domain model.}

@@

Pure syntax-based evaluation metrics can report low scores for learned models that are actually {\em sound} and {\em complete} but syntactically different from the GTM. Semantic evaluation metrics add a distinctive value over the syntactic ones, which is that they evaluate the learned model with a set of observations of plan executions. \textcolor{red}{These metrics measure how well a model can reproduce a given plan execution, so the use of the word "semantic" here is in reference to the degree to which the learned model is able to capture the physics of the domain.} Semantic metrics are appropriate for scenarios where:

\begin{enumerate}
	\item The GTM is unknown. This is the most common scenario in ML, where models are both learned and evaluated with respect to datasets.
%	\item The GTM is known but a test-based evaluation on a dataset is preferable (or needed as complementary to a syntactic evaluation). As a rule of thumb, it is preferable to evaluate the learned models wrt a dataset because a learned model can be semantically correct though syntactically incorrect (different from the GTM). We refer to this phenomenon as \emph{model reformulation}.
	\item \textcolor{red}{We are interested in measuring the ability of a model to explain a given plan trace, which is a good indicator of how the model will perform in actual planning tasks.} As a rule of thumb, it is preferable to evaluate the learned models wrt a dataset because a learned model can be semantically correct though syntactically incorrect (different from the GTM). We refer to this phenomenon as \emph{model reformulation}.
\end{enumerate}

@@

%These semantic alterations typically appear in the learned models when the observed input data given in $\tau$ is scarce. Defining a proper semantic evaluation is key because the application of syntax-based metrics may report low scores for learned models that are actually {\em sound} and {\em complete} but correspond to {\em reformulations} of the GTM. In the following, we introduce a novel evaluation metric that is robust to different types of reformulation.

\textcolor{red}{The \ARMS system was the first to show that a semantic evaluation can be done via validation of a set of plan traces with the learned model~\cite{yang2007learning}. The underlying idea is that an error indication of the learned action models is obtained by counting the number of preconditions that are not satisfied during the execution of the plan trace with the learned models, similarly to the functionality provided by the automatic validation tool VAL~\cite{howey2004val} used in the IPCs. This approach can be understood as modifying the plan trace (by adding the necessary preconditions to the intermediate states) so as to allow the execution of the observed actions using the learned models. In other words, modifying the plan trace to fit the model. Inspired by this approach, we present novel semantic-based error measure that builds upon the {\em precision} and {\em recall} metrics, and instead, determines the modifications required by a learned model to explain the given plan traces.}

%Definitions \ref{syn-precision} and \ref{syn-recall} are syntax-based metrics to evaluate $\mathcal{M}$ with respect to the GTM. A semantic metric, on the other hand, evaluates $\mathcal{M}$ with respect to a given set of plan traces.

@@

The intuition behind this evaluation is to {\em semantically} assess how well the learned domain model $\mathcal{M}$ explains a set of given observations of plan executions according to the amount of {\em edition} required by $\mathcal{M}$ to induce the observations. \textcolor{red}{The interpretation of a sound and complete model in the semantic perspective is slightly different from the syntactic one. In this case, a sound model is one were there is no need to remove any precondition or effect in order to be consistent with some given plan traces. A complete model is one that is consistent with the given plan traces without adding any of its preconditions and effects.} Unlike the semantic metric defined by ARMS, our novel semantic definitions of precision and recall are not sensitive to flaws in the action model that manifest more than once in the plan traces since the flaws are corrected only once in the learned models instead of at every intermediate state of the plan traces.

%This semantic evaluation approach is again flexible to various amount and kind of available input knowledge.

@@

%Our compilation is extensible to compute the {\em closest compliant set of action models} and hence, our semantic versions of the {\em precision} and {\em recall} metrics. This extension considers that the input models $\mathcal{M}$, is {\em non-empty} so instead of learning an action model from scratch we simply edit $\mathcal{M}$ until it satisfies the given input observations. In other words, now $\mathcal{M}$ is a set of given operator schemas, wherein each $\xi\in\mathcal{M}$ initially contains the $pre(\xi)$, $del(\xi)$ and $add(\xi)$ sets. A solution to the classical planning task resulting from the extended compilation is a sequence of actions that:

>>>>

In this section we introduce the metrics used by \FAMA to evaluate the action models that result from solving a learning task $\Lambda$. First, we will motivate the use of two standard syntactic metrics ({\em precision} and {\em recall}). Then, section \ref{syntactic_precision_recall} will define these metrics for planning models, and section \ref{semantic_precision_recall} will introduce a semantic evaluation measure that builds upon {\em precision} and {\em recall}. Finally, section \ref{edit_distance} explains how \FAMA computes our novel semantic-based metrics.

We can observe in Table \ref{table:models_comparison2} that most of the approaches use a similar syntax-based metric that consists in (1) counting the missing and extra fluents that appear in the learned model wrt the GTM and (2) normalizing this error by the the total number of all the possible preconditions and effects of an action model. This is an \emph{optimistic} metric since error rates are not normalized by the size of the actual GTM. The set of preconditions and effects of the GTM is usually smaller than the set of all possible preconditions and effects and thereby it turns out that these syntax-based metrics may output error rates below 100\% for totally wrong learned models. To overcome this limitation we propose to use two standard metrics from ML, {\em precision} and {\em recall}, that are frequently used in pattern recognition, information retrieval and binary classification~\cite{davis2006relationship}. These two syntactic metrics are generally more informative than counting the number of errors between the learned action models and the GTM:

@@

\begin{mydefinition}
	Let $PRE(\xi) = \underset{p \in pre(\xi)}{\bigcup} pre_p(\xi)$, $ADD(\xi) = \underset{p \in add(\xi)}{\bigcup} add_p(\xi)$, and $DEL(\xi) = \underset{p \in del(\xi)}{\bigcup} del_p(\xi)$ be the set of propositional fluents that represent preconditions, positive and negative effects of a given action model $\xi$. We define:
	\begin{small}

			\begin{align*}
			INS(\mathcal{M}, \mathcal{M'})=&\bigcup\limits_{\xi\in\mathcal{M}, \xi'\in\mathcal{M'} s.t.\ name(\xi) = name(\xi')}
			PRE(\xi') \backslash PRE(\xi) \cup
			ADD(\xi') \backslash ADD(\xi) \cup
			DEL(\xi') \backslash DEL(\xi)
			\\
			\\
			DEL(\mathcal{M}, \mathcal{M'})=&\bigcup\limits_{\xi\in\mathcal{M}, \xi'\in\mathcal{M'} s.t.\ name(\xi) = name(\xi')}
			PRE(\xi) \backslash PRE(\xi') \cup
			ADD(\xi) \backslash ADD(\xi') \cup
			DEL(\xi) \backslash DEL(\xi')
			\end{align*}

	\end{small}
\end{mydefinition}

With these ingredients in mind, we adapt the definitions of syntactic precision and recall to domain models. Let $\mathcal{M}$ be a domain model and let $\mathcal{M'}$ be the GTM. We know that $size(\mathcal{M}) = \sum_{\xi \in \mathcal{M}}\left|pre(\xi)\right| + \left|add(\xi)\right| + \left|del(\xi)\right|$ and by definition the number of preconditions and effects of the learned action models is equal to the sum of {\em true positives} and {\em false positives}; that is, $size(\mathcal{M}) = tp + fp$.

@@

\begin{mydefinition} \label{syn-precision} The precision of $\mathcal{M}$ relative to the GTM is defined as the fraction of the common preconditions and effects between $\mathcal{M}$ and the GTM among all prediconditions and effects of $\mathcal{M}$.
	\begin{small}
		\begin{align*}
		Precision=\frac{tp}{tp+fp}=\frac{size(\mathcal{M})- \left|DEL(\mathcal{M},GTM)\right|}{size(\mathcal{M})}
		\end{align*}
	\end{small}
\end{mydefinition}

\begin{mydefinition} \label{syn-recall} The recall of $\mathcal{M}$ relative to the GTM is defined as the fraction of the common preconditions and effects between $\mathcal{M}$ and the GTM among all preconditions and effects of the GTM.

@@

Intuitively, precision gives a notion of {\em soundness} while recall gives a notion of the {\em completeness} of the learned models. We interpret a sound learned model as one in which all preconditions and effects are correct with respect to the GTM, and so there is no need to remove anything. A complete model is one in which no precondition or effect is missing.
As an example, a precision of 0.5 means that only half of the predicates that make up the learned domain model are present in the GTM, while a recall of 0.5 means that only half of the predicates that make up the GTM are present in the learned domain model.

@@

Pure syntax-based evaluation metrics can report low scores for learned models that are actually {\em sound} and {\em complete} but syntactically different from the GTM. Semantic evaluation metrics add a distinctive value over the syntactic ones, which is that they evaluate the learned model with a set of observations of plan executions. These metrics measure how well a model can reproduce a given plan execution, so the use of the word "semantic" here is in reference to the degree to which the learned model is able to capture the physics of the domain. Semantic metrics are appropriate for scenarios where:

\begin{enumerate}
	\item The GTM is unknown. This is the most common scenario in ML, where models are both learned and evaluated with respect to datasets.
	\item We are interested in measuring the ability of a model to explain a given plan trace, which is a good indicator of how the model will perform in actual planning tasks. As a rule of thumb, it is preferable to evaluate the learned models wrt a dataset because a learned model can be semantically correct though syntactically incorrect (different from the GTM). We refer to this phenomenon as \emph{model reformulation}.
\end{enumerate}

@@

The \ARMS system was the first to show that a semantic evaluation can be done via validation of a set of plan traces with the learned model~\cite{yang2007learning}. The underlying idea is that an error indication of the learned action models is obtained by counting the number of preconditions that are not satisfied during the execution of the plan trace with the learned models, similarly to the functionality provided by the automatic validation tool VAL~\cite{howey2004val} used in the IPCs. This approach can be understood as modifying the plan trace (by adding the necessary preconditions to the intermediate states) so as to allow the execution of the observed actions using the learned models. In other words, modifying the plan trace to fit the model. Inspired by this approach, we present novel semantic-based error measure that builds upon the {\em precision} and {\em recall} metrics, and instead, determines the modifications required by a learned model to explain the given plan traces.

@@

The intuition behind this evaluation is to {\em semantically} assess how well the learned domain model $\mathcal{M}$ explains a set of given observations of plan executions according to the amount of {\em edition} required by $\mathcal{M}$ to induce the observations. The interpretation of a sound and complete model in the semantic perspective is slightly different from the syntactic one. In this case, a sound model is one were there is no need to remove any precondition or effect in order to be consistent with some given plan traces. A complete model is one that is consistent with the given plan traces without adding any of its preconditions and effects. Unlike the semantic metric defined by ARMS, our novel semantic definitions of precision and recall are not sensitive to flaws in the action model that manifest more than once in the plan traces since the flaws are corrected only once in the learned models instead of at every intermediate state of the plan traces.

@@
